From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-08-12 draft_v1/draft_v2.tex
vitor_feedback_2nd

\date{2025-07-26}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates how the internal structure of
depression--precariousness dynamics shapes responsiveness to external
support. Using cross-sectional data from the HELIUS study, we first
apply cycle-capable causal discovery algorithms to estimate directional
dependencies between financial stress, domain-specific precariousness,
and depressive symptoms. Our multi-resolution analysis---spanning both
aggregate and symptom-level constructs---highlights financial stress as
a plausible upstream driver of depression, with symptoms such as sleep
disturbance and depressed mood emerging as initiators of broader social
vulnerability. To translate these structural insights into dynamic
predictions, we construct a family of analytically calibrated linear
stochastic models. Simulations reveal that variations in internal
structure can substantially alter system responses to financial stress
reduction. Some models show rapid but transient improvements in
depression, while others adjust more slowly yet sustain gains over time.
Precariousness tends to respond more immediately across all
configurations, though its trajectory also shifts subtly with structural
variation. Together, these findings demonstrate that intervention
outcomes depend not only on the magnitude of external input but also on
latent properties of the system. More broadly, this study illustrates
how combining causal discovery with dynamical modeling supports
principled, simulation-based exploration of intervention
scenarios---grounded in data, yet able to probe counterfactual
structure. This approach offers a scalable framework for diagnosing
hidden system constraints and informing targeted public health
strategies in complex psychosocial domains.
\end{abstract}

@@

The sum score graphs provide a high-level summary of how precariousness
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precariousness factors (\emph{P.hou},
\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
depression sum score (\emph{PHQsum}), as identified by the FCI
algorithm. As described in Section~\ref{sec-analysis}, dashed edges
indicate ties, and the edge endpoint at which the tie occurs is marked
by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the
relationships between precariousness factors and depression
(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
precariousness (\emph{P.soc}) are not identified as causes of
depression, whereas financial stress (\emph{S.fin}) appears to play a
causal role. While \emph{P.emp} and \emph{P.soc} are generally not
recognized as causes of other precariousness factors, \emph{S.fin}
emerges as a potential cause, as indicated by its circle edge endpoint.
This pattern is further supported by the edge-type proportion matrix
shown in panel (b), which aggregates edge inference outcomes across
bootstrapped datasets.

The matrix plot can be interpreted such that the symbol in
matrix{[}\(i\), \(j\){]} represents the relationship
\(i -~[symbol] \  j\). For example, if matrix{[}\(i\), \(j\){]} is
\textgreater{} and matrix{[}j, i{]} is \(\circ\), then the inferred
relationship is \(i \circarrow j\). In the table, different edge types
are represented by distinct colors: light gray for the absence of an
edge, blue for circles, green for arrowheads, and coral for arrowtails.
The colors are blended based on the proportion of each edge type, with
higher proportions increasing the opacity of the corresponding color,
making dominant symbols more visually prominent.

@@

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI algorithm, with one
exception: CCI does not identify \emph{S.fin} as a cause of depression.
Additionally, CCI introduces greater uncertainty in edge directions,
with more ties mainly between circle and arrowhead endpoints, and tends
to favor arrowheads more frequently than FCI. Despite these differences,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm, supporting the key
pathways.

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI analysis, with one key
distinction: CCI shows greater ambiguity in the directional relationship
between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
remaining cases display an undirected (circle) endpoint. Overall, CCI
introduces greater uncertainty in edge orientations relative to FCI,
yielding more undirected endpoints and more frequent bidirectional
(confounded) edges, which in turn contribute to a slightly higher
overall count of arrowheads when directions are inferred. Despite this,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm. For example, both
algorithms identify directional pathways from the stressors
(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc}, and suggest that
\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
by it. \emph{P.emp} is not inferred to cause any of its neighboring
variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}), and instead
appears to be affected by them. \emph{P.hou}, in turn, is connected to
\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
may be a potential source of influence rather than an outcome. These
consistent pathways underscore the robustness of the core relational
architecture across algorithms amid variation in directional confidence.

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure, the symptom-level graph reveals the heterogeneity in how
specific symptoms (\emph{con} (concentration), \emph{slp} (sleep),
\emph{ene} (energy), \emph{app} (appetite), \emph{mot} (motor),
\emph{sui} (suicidal), \emph{anh} (anhedonia), \emph{glt} (guilt), and
\emph{dep} (depressed mood)) are influenced by, and in turn influence,
different forms of precariousness.

The symptom-level graph in Figure~\ref{fig-sym-1} reveals a more complex
and interconnected structure than the sum score graph, reflecting the
strong interdependence among symptoms and suggesting much presence of
latent confounders, as indicated by numerous bidirectional edges. As
before, dashed edges denote ties, with specific ties marked by bold
horizontal bars. In Figure~\ref{fig-sym-2}, the proportion matrix
provides more insight into these directionalities. Here, opaque green
dominates most symptom-to-symptom connections, indicating that
arrowheads are the most frequently inferred edge type. However, blue
regions, which correspond to circle endpoints, are particularly common
for \emph{anh}, \emph{slp}, \emph{ene}, and \emph{sui}, reflecting
uncertainty in the causal direction among these symptoms. Notably, the
only arrowtail connection appears between \emph{dep} and \emph{anh},
suggesting that anhedonia is a potential cause of depressed mood.

@@

This discrepancy likely arises from a combination of factors. As the
analysis moves from aggregated to disaggregated variables, the number of
CI tests increases substantially, leading to a loss of statistical
power. With more variables being conditioned on, the effective sample
size per test diminishes, making it harder to detect subtle
dependencies. Over-conditioning---controlling for too many
variables---can also obscure real relationships, resulting in false
negatives. Beyond these statistical power issues, the underlying
assumptions of the two CI tests contribute to their divergence. RCoT is
a nonparametric test that captures both linear and nonlinear
dependencies without assuming a specific distributional form. However,
it is more conservative and requires larger sample sizes, especially in
high-dimensional settings. In contrast, the Gaussian CI test is based on
partial correlations and assumes linear Gaussian relationships, making
it more permissive and more likely to detect weak or spurious
associations. As a result, it is common to see edges where Gaussian CI
suggests a directional relationship, while RCoT indicates no edge. In
the proportion matrix, this appears as a 50--50 tie between absence and
arrowhead symbols.

@@

Other precariousness factors exhibit more uncertainty but still notable
relationships. \emph{S.rel} shows weak but existing connections with
\emph{slp}, \emph{glt}, and \emph{app}, though directionality remains
unclear in many cases. \emph{P.emp} also connects to symptoms like
\emph{slp}, \emph{mot}, and \emph{con}, but, like \emph{S.rel}, these
relationships exhibit a 50/50 split in directionality, reflecting
uncertainty in the inferred causal paths. In line with the aggregated
graph in Figure~\ref{fig-sum}, \emph{P.hou} does not appear to have any
direct relationship with depression symptoms but consistently shows
associations with \emph{P.emp} and \emph{S.fin}. The causal
relationships among precariousness factors remain largely unchanged from
the aggregated analysis, with \emph{S.fin} exhibiting the strongest
tendency to influence other precariousness factors.

Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
to be the most connected to precariousness factors, while \emph{P.soc}
has the most connections with symptoms, predominantly as a recipient
rather than a driver of influence. Within the symptom network,
\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
symptoms, as they exhibit more outgoing arrows compared to other
symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
\emph{glt}, despite having high connectivity, predominantly receive
incoming arrows, indicating they are more likely effects rather than
causes. Considering both symptom-precariousness connections and
symptom-level dynamics, \emph{slp} emerges as a central symptom, given
its strong ties to precariousness factors and its influential role
within the symptom network. This suggests that sleep issues could be an
initiating symptom, particularly sensitive to relational stress and
employment precariousness. On the other hand, \emph{P.soc} serves as a
key bridge between the depression and precariousness subsystems, as
depressive symptoms appear to feed back into social precariousness,
reinforcing a self-sustaining dynamic between depression and precarious
conditions.

@@

Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows more
certainty in causal directions, with only a few edges resulting in ties,
all of which involve the overall precariousness factor. This is even
more evident in the proportion matrix (Figure~\ref{fig-presum-2}), where
symptom-to-symptom interactions almost fully converge to a single edge
type. Additionally, all symptom interactions are connected either
through bidirectional edges or a combination of an arrowhead and a
circle, suggesting a strong presence of latent confounders.

Examining the symptom-precariousness connections, we observe a stronger
overall trend of depressive symptoms influencing precariousness, rather
than the other way around. Specifically, \emph{anh}, \emph{dep}, and
\emph{slp} show a high probability of causing precariousness, whereas
\emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty in
directionality, often resulting in circle endpoints. This pattern
suggests that when precariousness factors are aggregated, the dominant
causal flow is from depressive symptoms to precariousness, rather than
precariousness driving depression. Particularly, \emph{dep} emerges as
the strongest predictor of precariousness, exhibiting the highest
proportion of arrowtails, consistent with findings from the
disaggregated analysis.

@@

Overall, the aggregated precariousness graph supports the key findings
from both the disaggregated analysis and the depression sum score
analysis, providing greater certainty in several important patterns. The
results highlight the significant role of latent confounders in
symptom-to-symptom interactions. While certain stressor-related
precariousness factors, particularly \emph{S.fin} and \emph{S.rel},
appear to causally influence depressive symptoms, other precariousness
factors are more likely driven by depression rather than acting as
primary causes, suggesting that precariousness may function more as a
consequence of depression than its source. Some symptoms, particularly
\emph{glt}, \emph{slp}, and \emph{dep}, appear to influence precarious
conditions while also occupying central roles within the symptom
network. Among them, \emph{slp} and \emph{glt} emerge as influential
symptoms within the depression network, whereas \emph{con} and
\emph{dep}, despite their high connectivity, behave more as outcomes
rather than primary drivers. Together, these findings suggest that
depression---especially symptoms like sleep disturbances and guilt---may
contribute to sustaining precarious conditions, possibly forming a
reciprocal cycle between depression and precariousness.

@@

\caption{\label{fig-presum}Resulting graph of precariousness sum score
and individual depression symptoms using FCI and proportion of edge
endpoint types.}

@@

To assess how internal structure influences the system's responsiveness
to external support, we simulate a set of linear models that are
consistent with the data but vary in the assumed strength of influence
from precariousness to depression (\(\alpha_{DP}\)). For each model, we
generate a virtual population of 1,000 individuals and simulate an
external intervention that reduces financial stress (\(S\)), which acts
as a shared input across individuals. This intervention defines the
control input, while different values of \(\alpha_{DP}\) represent
alternative internal feedback scenarios. We examine system behavior in
two complementary ways: (1) by analyzing average levels of depression
and precariousness at selected time points following intervention onset,
and (2) by simulating full temporal trajectories under sustained
intervention followed by recovery, using two representative cases at the
extremes of the \(\alpha_{DP}\) range.

Figure~\ref{fig-intervention} summarizes how variation in
\(\alpha_{DP}\) shapes system-level responses. Panel A shows the mean
levels of depression (\(D\)) and precariousness (\(P\)) at time steps
100, 300, and 500, with the x-axis indicating the strength of external
support, defined as the proportional reduction of \(S\) toward its
minimum observed value. Each line represents a different scenario with a
different \(\alpha_{DP}\), color-coded from low (purple) to high
(yellow).

In the early phase of intervention (\(t = 100\)), depression and
precariousness exhibit contrasting patterns. Depression declines more
rapidly in scenarios with lower \(\alpha_{DP}\), where its dynamics are
driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
increases, \(D\) becomes more dependent on precariousness (\(P\)), which
itself adjusts more slowly---delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in
scenarios with higher \(\alpha_{DP}\). This reflects a shift in
influence: as \(\alpha_{DP}\) increases, \(P\) becomes more directly
responsive to \(S\) (via increasing \(\alpha_{PS}\)) and less influenced
by \(D\), allowing it to adjust more promptly to changes in the
stressor.

Despite these short-term differences, all scenarios ultimately converge
to similar steady-state levels under sustained intervention. By time
step 300, convergence is already substantial---though some variation
remains, particularly in \(P\), where gradients between models are still
visible. By step 500, trajectories have nearly fully aligned across all
configurations, consistent with the shared equilibrium structure of the
linear system. This convergence highlights that the key distinctions lie
not in whether intervention is effective, but in how quickly and through
which pathways change unfolds.

Panel B shows full time trajectories for two representative cases: one
with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, intervention
begins at step 300 and ends at step 1500. When \(\alpha_{DP} = 0\),
depression responds rapidly to the reduction in stress and rebounds
shortly after support is withdrawn. In contrast, when \(\alpha_{DP}\) is
high, \(D\) responds more gradually and recovers more
slowly---reflecting its increased dependence on the slower-changing
\(P\). Precariousness declines in both scenarios during intervention and
returns to baseline afterward, but with a slightly sharper drop in the
higher \(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
directly responsive to \(S\), as \(\alpha_{PS}\) increases with
\(\alpha_{DP}\).

@@

\caption{\label{fig-intervention}System response to external
intervention across variations in internal structure.
\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
precariousness (\(P\)) at time steps 100, 300, and 500, across different
values of \(\alpha_{DP}\) and intervention strengths. Models with lower
\(\alpha_{DP}\) show faster early reductions in \(D\), while higher
\(\alpha_{DP}\) leads to more immediate declines in \(P\). \textbf{Panel
B}: Full time trajectories for two models---one with no coupling
(\(\alpha_{DP} = 0\)) and one with strong coupling
(\(\alpha_{DP} = 0.65\))---illustrating system response and recovery
following intervention onset (step 300) and withdrawal (step 1500).
Depression responds more gradually and persistently in the high
\(\alpha_{DP}\) model, whereas precariousness adjusts more quickly.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact---and how their internal structure modulates responsiveness to
intervention. Using cross-sectional data from the HELIUS study, we
applied cycle-capable discovery algorithms to estimate plausible
directional relationships among multiple domains of precariousness
factors and depressive symptoms. By analyzing both aggregate constructs
and individual variables, we aimed to capture patterns that may be
masked in either extreme of abstraction (i.e., purely aggregate or
disaggregated analyses).

The causal analysis consistently identified financial stress as a key
plausible driver of depression, appearing across both aggregated and
disaggregated models. At the symptom level, several nodes---sleep
disturbance, depressed mood, guilt, and anhedonia---emerged as
especially sensitive to external influences. Among these, depressed mood
played a particularly central role: it was influenced by other symptoms
and, in turn, showed directional influence on social precariousness,
suggesting a feedback mechanism wherein worsening mood fosters social
withdrawal, which may subsequently exacerbate depressive symptoms. The
CCI algorithm further suggested a potential cyclic link between
depressed mood and overall precariousness, hinting that symptom and
precariousness factors may co-evolve in mutually reinforcing ways.
Symptom sleep disturbance also played a key initiating role. It showed
directed connections to both social and employment precariousness and
exhibited many outgoing edges within the symptom network. This suggests
the symptom sleep disturbance may function as an early warning signal or
entry point into the depression--precariousness cycle, making it a
promising target for preventive intervention. In contrast, social
precariousness more often appeared as an effect rather than a
cause---receiving arrows from symptoms like depressed mood and guilt,
but rarely sending them. This asymmetry implies that social
precariousness may often reflect downstream consequences of worsening
mental health rather than being its root cause (Compton \& Shim, 2015;
Lund et al., 2010). While such relationships are complex and likely
reciprocal over time, the directional patterns observed provide a useful
foundation for hypothesis generation and targeted intervention.

Taken together, these patterns offer actionable insights for
intervention. Symptoms such as sleep disturbance, which show early and
wide-reaching influence within the symptom network, may serve as
effective targets for prevention---limiting escalation into broader
depressive states or worsening precariousness. Similarly, depressed
mood, which functions as a central hub influencing both symptom and
precariousness variables, may offer a leverage point for disrupting
reinforcing cycles of mental health decline and social isolation. At the
same time, the broader causal structure points to interconnected
feedback loops that span both psychological symptoms and external
conditions such as financial and social precariousness---factors that
often reflect the influence of structural inequalities (e.g., income
instability, housing insecurity, or social marginalization). These
complex dynamics suggest that single-point interventions may be
insufficient to produce lasting change. Supporting this, recent
simulation work by Park, Li, et al. (2025) demonstrates that feedback
loops---even when modest in strength---can substantially reduce the
long-term impact of isolated interventions, as system dynamics tend to
re-amplify vulnerability through alternate pathways. In line with
emerging perspectives in systems-oriented mental health research
(Borsboom, 2017), our findings underscore the need for multi-pronged
strategies that address not only internal symptom dynamics but also the
broader socioeconomic conditions that sustain vulnerability. Intervening
simultaneously at multiple points of reinforcement may be essential to
shift the system out of a maladaptive attractor and toward more stable
trajectories of well-being.

>>>>

\date{2025-08-12}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates how the internal structure of
depression--precariousness dynamics shapes responsiveness to external
support. Using cross-sectional data from the HELIUS study, we first
apply cycle-capable causal discovery algorithms to estimate directional
dependencies between financial stress, domain-specific precariousness,
and depressive symptoms. Our multi-resolution analysis---spanning both
aggregate and symptom-level constructs---highlights financial stress as
a plausible upstream driver of depression, with symptoms such as sleep
disturbance and depressed mood emerging as initiators of broader social
vulnerability. To translate these structural insights into dynamic
predictions, we construct a family of analytically calibrated linear
stochastic models. Simulations reveal that variations in internal
structure can substantially alter system responses to financial stress
reduction. Some models show rapid but transient improvements in
depression, while others adjust more slowly yet sustain gains over time.
Precariousness tends to respond more immediately across all
configurations, though its trajectory also shifts subtly with structural
variation. Together, these findings demonstrate that intervention
outcomes depend not only on the magnitude of external input but also on
latent properties of the system. More broadly, this study illustrates
how combining causal discovery with dynamical modeling supports
principled, simulation-based exploration of intervention
scenarios---grounded in data, yet able to probe counterfactual
structure. This approach offers a scalable framework for diagnosing
hidden system constraints and informing targeted public health
strategies in complex psychosocial domains.
\end{abstract}

@@

Figure~\ref{fig-sum} provides a high-level summary of how precariousness
factors collectively influence and are influenced by overall depression
severity, focusing on the causal relationships between precariousness
factors (\emph{P.hou}, \emph{P.emp}, \emph{P.soc}, \emph{S.rel},
\emph{S.fin}) and the depression sum score (\emph{PHQsum}), as
identified by the FCI algorithm.

In panel (a) we see that employment precariousness (\emph{P.emp}) and
social precariousness (\emph{P.soc}) are not identified as causes of
depression, whereas financial stress (\emph{S.fin}) appears to play a
causal role. While \emph{P.emp} and \emph{P.soc} are generally not
recognized as causes of other precariousness factors, \emph{S.fin}
emerges as a potential cause, as indicated by its circle edge endpoint.
This pattern is further supported by the edge-type proportion matrix
shown in panel (b), which aggregates edge inference outcomes across
bootstrapped datasets.

@@

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI analysis, with one key
distinction: CCI shows greater ambiguity in the directional relationship
between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
remaining cases display an undirected (circle) endpoint. Overall, CCI
introduces greater uncertainty in edge orientations relative to FCI,
yielding more undirected endpoints and more frequent bidirectional
(confounded) edges, which in turn contribute to a slightly higher
overall count of arrowheads when directions are inferred. Both
algorithms identify directional pathways from the stressors
(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc} and suggest that
\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
by it. \emph{P.emp} is not inferred to cause any of its neighboring
variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}) and instead
appears to be affected by them. \emph{P.hou}, in turn, is connected to
\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
may be a potential source of influence rather than an outcome.

Causal relationships between five precariousness factors and depression
sum score (PHQsum), as estimated using the FCI algorithm across
bootstrapped samples. (a) Aggregated causal graph showing the most
frequent edge types across bootstraps. Dashed edges indicate ties, with
a bold horizontal bar marking where the tie occurs. (b) Proportion
matrix summarizing edge types across all bootstraps. Each matrix cell
\([i, j]\) represents how often an edge from variable \(i\) to variable
\(j\) exhibited a particular edge mark---arrowhead, tail, circle, or no
edge---adjacent to \(j\). The edge mark in matrix\([i,j]\) reflects the
orientation of the edge ending at \(j\). For example, if
matrix\([i, j]\) shows a green arrowhead (\(>\)) and matrix\([j, i]\)
shows a blue circle (\(\circ\)), then the inferred relationship is
\(i \circarrow j\). Edge types are encoded by distinct colors: green for
arrowheads, coral for arrowtails, blue for circles, and light gray for
absence of an edge. Colors are blended and shaded by frequency, with
greater opacity indicating higher confidence in the inferred edge type.

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure, the symptom-level graph in Figure~\ref{fig-sym} reveals the
heterogeneity in how specific symptoms (\emph{con} (concentration),
\emph{slp} (sleep), \emph{ene} (energy), \emph{app} (appetite),
\emph{mot} (motor), \emph{sui} (suicidal), \emph{anh} (anhedonia),
\emph{glt} (guilt), and \emph{dep} (depressed mood)) are influenced by,
and in turn influence, different forms of precariousness.

Figure~\ref{fig-sym-1} reveals a strong interdependence among symptoms
and suggests much presence of latent confounders, as indicated by
numerous bidirectional edges. In Figure~\ref{fig-sym-2}, the proportion
matrix indicates that arrowheads are the most frequently inferred edge
type. However, circle endpoints are particularly common for \emph{anh},
\emph{slp}, \emph{ene}, and \emph{sui}, reflecting uncertainty in the
causal direction among these symptoms. Notably, the only arrowtail
connection appears between \emph{dep} and \emph{anh}, suggesting that
anhedonia is a potential cause of depressed mood.

@@

@@

Other precariousness factors exhibit more uncertainty but still notable
relationships. \emph{S.rel} is connected to \emph{slp}, \emph{glt}, and
\emph{app}, though directionality remains unclear in many cases.
\emph{P.emp} also connects to symptoms like \emph{slp}, \emph{mot}, and
\emph{con}, but, like \emph{S.rel}, these relationships exhibit a 50/50
split in directionality, reflecting uncertainty in the inferred causal
paths. In line with the aggregated graph in Figure~\ref{fig-sum},
\emph{P.hou} does not appear to have any direct relationship with
depression symptoms but consistently shows associations with
\emph{P.emp} and \emph{S.fin}. The causal relationships among
precariousness factors remain largely unchanged from the aggregated
analysis, with \emph{S.fin} exhibiting the strongest tendency to
influence other precariousness factors.

Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
to be the most connected to precariousness factors, while \emph{P.soc}
has the most connections with symptoms, predominantly as a recipient
rather than a driver of influence. Within the symptom network,
\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
symptoms, as they exhibit more outgoing arrows compared to other
symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
\emph{glt}, despite having high connectivity, predominantly receive
incoming arrows, indicating they are more likely effects rather than
causes. Considering both symptom-precariousness connections and
symptom-level dynamics, \emph{slp} emerges as a central symptom, given
its strong ties to precariousness factors and its influential role
within the symptom network. \emph{P.soc} appears to function as a bridge
between the depression and precariousness subsystems, as depressive
symptoms appear to feed back into social precariousness, reinforcing a
self-sustaining dynamic between depression and precarious conditions.

@@

Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows greater
certainty in causal directions, with ties occurring only in edges
involving the overall precariousness factor. This is even more evident
in the proportion matrix (Figure~\ref{fig-presum-2}), where
symptom-to-symptom interactions almost fully converge to a single edge
type. Additionally, all symptom interactions are connected either
through bidirectional edges or a combination of an arrowhead and a
circle, suggesting a strong presence of latent confounders.

Examining the symptom-precariousness connections, we observe a stronger
overall trend of depressive symptoms influencing precariousness, rather
than the other way around. Specifically, \emph{anh}, \emph{dep}, and
\emph{slp} frequently emerge as sources of influence on precariousness,
whereas \emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty
in directionality, often resulting in circle endpoints. This pattern
suggests that when precariousness factors are aggregated, the dominant
causal flow is from depressive symptoms to precariousness, rather than
precariousness driving depression. Particularly, \emph{dep} emerges as
the strongest predictor of precariousness, exhibiting the highest
proportion of arrowtails, consistent with findings from the
disaggregated analysis.

@@

Across multiple levels of analysis, our results point to a consistent
and directional link between precariousness and depression. Financial
stress (\emph{S.fin}) emerges as a particularly robust and modifiable
upstream driver: in both aggregated and disaggregated models, it shows
strong and stable causal influence on depressive symptoms. By contrast,
other forms of precariousness---particularly social precarity---more
often appear as outcomes rather than antecedents of depression,
frequently receiving rather than sending causal arrows. Among symptoms,
\emph{dep} and \emph{slp} stand out for their centrality and connections
to precariousness factors, with \emph{dep} also implicated in potential
feedback loops with social precarity (\emph{P.soc}). These findings
motivate the use of a simplified three-node model in the
simulation---capturing depression, financial stress, and social
precarity---as it preserves the key dynamic motifs observed in the
causal graphs while enabling tractable exploration of system-level
responses to stress-reducing interventions.

@@

\caption{\label{fig-presum}Resulting graph of precariousness sum score
and individual depression symptoms using FCI and proportion of edge
endpoint types. As before, dashed edges denote ties, with specific ties
marked by bold horizontal bars.}

@@

To assess how internal structure influences the system's responsiveness
to external support, we simulate a set of linear models that are
consistent with the data but vary in the assumed strength of influence
from precariousness to depression (\(\alpha_{DP}\)). For each model, we
generate a virtual population of 1,000 individuals and simulate an
external intervention that reduces each individual's financial stress
(\(S\)) proportionally toward the minimum observed value in the dataset.
While individuals begin with different baseline stress levels, the
intervention applies a shared proportional reduction across the
population. This intervention defines the external input, while
variation in \(\alpha_{DP}\) reflects different assumptions about
internal feedback from precariousness to depression. We examine system
behavior in two complementary ways: (1) by analyzing average levels of
depression and precariousness at selected time points following
intervention onset, and (2) by simulating full temporal trajectories
under sustained intervention followed by recovery, using two
representative cases at the extremes of the \(\alpha_{DP}\) range.

Figure~\ref{fig-intervention} summarizes how variation in
\(\alpha_{DP}\) shapes system-level responses to external support. In
Panel A, during the early phase of intervention, depression and
precariousness exhibit contrasting patterns. Depression declines more
rapidly in scenarios with lower \(\alpha_{DP}\), where its dynamics are
driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
increases, \(D\) becomes more dependent on precariousness (\(P\)), which
itself adjusts more slowly---delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in
scenarios with higher \(\alpha_{DP}\). This reflects a shift in
influence: as \(\alpha_{DP}\) increases, \(P\) becomes more directly
responsive to \(S\) (via increasing \(\alpha_{PS}\)) and less influenced
by \(D\), allowing it to adjust more promptly to changes in the
stressor.

Despite these short-term differences, all scenarios ultimately converge
to similar steady-state levels under sustained intervention. AS the
systm approaches convergence, some variation remains, particularly in
\(P\), where gradients between models are still visible. This
convergence highlights that the key distinctions lie not in whether
intervention is effective, but in how quickly and through which pathways
change unfolds.

Panel B shows full time trajectories for two representative cases: one
with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, there is a
temporary intervention period. When \(\alpha_{DP} = 0\), depression
responds rapidly to the reduction in stress and rebounds shortly after
support is withdrawn. In contrast, when \(\alpha_{DP}\) is high, \(D\)
responds more gradually and recovers more slowly---reflecting its
increased dependence on the slower-changing \(P\). Precariousness
declines in both scenarios during intervention and returns to baseline
afterward, but with a slightly sharper drop in the higher
\(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
directly responsive to \(S\), as consistency with the data requires
\(\alpha_{PS}\) to increase with \(\alpha_{DP}\) (see
Figure~\ref{fig-tradeoff}). These parameter shifts reflect trade-offs
needed to preserve the empirical covariances, with further details
provided in Appendix Section~\ref{sec-analcalibration}.

@@

\caption{\label{fig-intervention}System response to external
intervention across variations in internal structure.
\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
precariousness (\(P\)) at time steps 100, 300, and 500, across varying
values of \(\alpha_{DP}\) and intervention strengths. The x-axis
represents the strength of external support, operationalized as the
proportional reduction of financial stress (\(S\)) toward its minimum
observed value. Each line corresponds to a different model defined by a
specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
and recovery following intervention onset (step 300) and withdrawal
(step 1500). Depression responds more gradually and persistently in the
high \(\alpha_{DP}\) model, whereas precariousness adjusts more
quickly.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact and how their internal structure modulates responsiveness to
intervention. Consistent with prior work linking financial strain to
psychological distress (Ezzy, 1993; Lund et al., 2010), our causal
analysis identified financial stress as a key upstream driver of
depression---emerging across both aggregated and disaggregated models.
This provides empirical support for longstanding hypotheses that
economic stressors can act as modifiable entry points into broader
cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of
psychopathology but offers a more fine-grained map of potential causal
relationships, generating concrete, testable hypotheses. It also points
to possible limitations of prevailing models, as some observed
symptom--symptom links may be shaped by shared confounders rather than
direct causal influence. This suggests that, in some contexts, broader
mechanisms or clusters of symptoms may provide more effective units of
analysis and intervention than individual symptoms alone. Within this
framework, sleep disturbance emerged as a potentially initiating
symptom, with directed connections to both symptom-level and
precariousness variables---mirroring longitudinal findings that identify
sleep disruption as an early warning signal for depressive episodes and
social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
Depressed mood, another core symptom, occupied a central position in the
network. Not only was it influenced by upstream symptoms like anhedonia
and guilt, but it also showed directional influence on social
precariousness---suggesting that internal psychological states may
actively shape external vulnerability. The CCI algorithm further
identified a potential feedback loop between depressed mood and overall
precariousness, resonating with social drift theories in depression,
which posit that declining mental health can lead to social exclusion,
job instability, and housing insecurity, reinforcing depressive cycles
(Boschloo et al., 2015; Compton \& Shim, 2015).

In contrast, social precariousness itself appeared more often as an
effect than a cause, receiving arrows from symptoms like depressed mood
and guilt but rarely sending them. This asymmetry supports prior
observations that indicators of social vulnerability often reflect
downstream consequences of poor mental health, though the relationships
are likely bidirectional over time (Lund et al., 2010). These patterns,
while derived from Amsterdam-specific data, align with mental health
research in other high-income city settings, where structural
inequalities such as poverty, unemployment, and housing insecurity are
persistent drivers of poor psychological outcomes (Fryers et al., 2003;
Weich \& Lewis, 1998). Although the specific dynamics may vary by
context, the parallels with broader literature suggest that similar
mechanisms could operate in comparable settings. Nonetheless,
replication in other populations will be important for assessing
robustness and external validity.

Taken together, these findings yield actionable insights for mental
health intervention. Symptoms like sleep disturbance, which demonstrate
early and broad influence across both symptom and precariousness
networks, may serve as effective targets for prevention---potentially
halting the progression into more entrenched depressive states or
worsening precariousness. Similarly, intervening on central nodes like
depressed mood may offer strategic leverage for breaking mutually
reinforcing cycles of mental decline and social isolation. Importantly,
interventions targeting such nodes may have ripple effects across
domains. For instance, evidence suggests that treating core depressive
symptoms can improve social functioning, employment engagement, and
housing stability---particularly when combined with social or financial
supports (Lund et al., 2018; McDaid et al., 2019). These dual benefits
underscore the value of integrated, cross-domain strategies that
simultaneously address internal symptoms and external stressors.
